---
title: "Reading direction from methylation: a first DMSA analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading direction from methylation: a first DMSA analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dmsa)
```

## The problem this package exists to solve

A CpG in a promoter and a CpG in a gene body carry opposite consequences for
expression when their methylation moves the same way. Conventional set-level
methods average them, so a gene whose probes move *coherently on the expression
scale* can cancel itself to nothing on the methylation scale.

This is not a power problem that a larger sample fixes. At a 50/50 composition
the cancellation is exact and a direction-blind test is null by construction.

We can build exactly that situation and watch it happen.

```{r simulate}
set.seed(1)

n <- 300L                      # participants
K <- 12L                       # probes in the gene
f <- rnorm(n)                  # the latent process the gene actually tracks

# Half the probes are predicted to RAISE expression when methylated (d = +1),
# half to LOWER it (d = -1). On the expression scale they agree. On the
# methylation scale they point in opposite directions.
d <- rep(c(1, -1), each = K / 2)

M <- sapply(seq_len(K), function(j) d[j] * 0.55 * f + sqrt(1 - 0.55^2) * rnorm(n))
colnames(M) <- sprintf("cg%04d", seq_len(K))

dat <- data.frame(y = 0.35 * f + rnorm(n), cv = rnorm(n))
```

## What a direction-blind test sees

Average the probes and test the average. This is, in essence, what a
composite-score method does when it ignores direction.

```{r blind}
blind <- summary(lm(rowMeans(M) ~ y + cv, data = dat))$coefficients["y", ]
round(blind, 4)
```

The estimate sits near zero and the *p*-value near 1. The signal did not fail to
exist; it was destroyed in the averaging step.

## The alignment step

`dmsa_align()` takes the per-probe direction calls and turns them into a
multiplier. `d` is the methylation-to-expression direction, `p_plus` the
confidence that the direction is positive. At the gene level no polarity is
applied; at the system level each gene is additionally signed against its
system's activation tone.

```{r align}
direction <- data.frame(
  cpg    = colnames(M),
  d      = d,
  p_plus = ifelse(d > 0, 0.9, 0.1)
)

al <- dmsa_align(direction, genes = rep("GENE1", K), level = "gene")
al[, c("probe", "d", "s", "p_s_plus", "usable")]
```

`s` is the sign each probe contributes once it is read on the expression scale.
Every probe now votes the same way, which is what the biology said all along.

## Running the three lenses

`dmsa_triangulate()` tests the unit through three pre-specified lenses on one
shared permutation stream: **coherence** (pooled aligned burden, which rewards
cross-probe sign concordance), **composite** (the unit's aligned score as a
single regressor), and **diffuse** (a confidence-weighted quadratic score). An
ACAT omnibus combines them, and because ACAT is valid under arbitrary
dependence, hedging across the three costs no multiplicity.

```{r triangulate}
res <- dmsa_triangulate(
  M         = M,
  data      = dat,
  rhs       = c("y", "cv"),
  term      = "y",
  units     = rep("GENE1", K),
  alignment = al,
  B         = 499,
  seed      = 1
)

res[, c("unit", "n_probes", "concordance", "direction",
        "p_coherence", "p_composite", "p_omnibus")]
```

The same data that read as null a moment ago now carry the signal. Nothing about
the measurements changed — only whether the aggregation step was told which way
each probe points. `concordance` of 1.00 says every usable probe agreed once
aligned, and `direction` gives the sign of the effect on the expression scale.

Permutation *p*-values have granularity `1/(B + 1)`, so at `B = 499` the
smallest attainable value is 0.002 and the numbers above are at that floor.
Raise `B` for a finer resolution; the manuscript uses `B = 1999`.

## Which lens fires is itself a result

The three lenses fail in different ways, so their pattern is informative rather
than redundant. Coherence rewards a unit whose probes agree in sign; composite
rewards a unified shift that survives as one regressor; diffuse rewards signal
spread across many probes whichever way each points. A finding that fires on
coherence alone has a different shape from one that fires on composite alone,
and the package reports both rather than collapsing them to a single number.

## What DMSA does not do

Directional alignment buys power against cancellation and costs power elsewhere.
Where the question is whether a *whole panel* is associated with an outcome, and
no directional structure is claimed, a quadratic score such as `globaltest`
detects associations that DMSA's pooled score does not — in the manuscript's
benchmark, that comparison is one DMSA loses. Alignment is a bet that the
direction calls carry information. When they do not, the bet costs something.

## Moving to real data

Two things replace the simulation above:

- **Direction calls.** `d` and `p_plus` come from a curated methylation-to-expression
  map. The companion package `cpgdirection`
  (<https://github.com/teindor/cpgdirection>, archived at
  <https://doi.org/10.5281/zenodo.22024185>) supplies them for 3.1 million
  CpG-gene pairs across four evidence classes, but any table with the same three
  columns works. `dmsa` does not require it.
- **The two-call interface.** `dmsa_frame()` declares the design once — outcome,
  covariates, permutation blocks, the level hierarchy — and `dmsa_report()` runs
  it and emits figures, machine-readable tables at every level, and a written
  paragraph stating what was tested and what was controlled. See
  `?dmsa_frame` for the full argument set.

```{r session}
sessionInfo()
```
