---
title: "One door in: from any preprocessing pipeline to DMSA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{One door in: from any preprocessing pipeline to DMSA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

DMSA models probes as the multivariate response, so every engine wants the
same two objects: a **samples x probes** matrix and a phenotype `data.frame`
with rows in the same order. No mainstream preprocessing pipeline ends there.
minfi ends in a `GenomicRatioSet` (probes in rows), sesame in a beta matrix
(probes in rows), meffil and ewastools in matrices (probes in rows), ChAMP in
`myNorm` plus a `pd` sheet, RnBeads in an `RnBSet`, methylprep in a CSV
export. The gap between those shapes and the analysis is where silent
mistakes live: a transposed matrix, M-values treated as betas, a sample sheet
in a different order than the arrays.

`dmsa_import()` is the one door in. It takes whichever end object a pipeline
produced and returns an object `dmsa_frame()` accepts directly:

```r
dmsa_frame(dmsa_import(x, pheno = sheet), outcome = ..., ...)
```

(For a curated `SummarizedExperiment` where you want to pick the assay
yourself - a `recountmethylation` download, say - the lower-level
`dmsa_se()` does the split with no further opinions.)

## The adapter in one example

```{r adapter-demo}
library(dmsa)
set.seed(1)
## the shape most pipelines produce: probes in rows, samples in columns
B <- matrix(runif(60), nrow = 6,
            dimnames = list(sprintf("cg%08d", 1:6), paste0("S", 1:10)))
ph <- data.frame(Sample_Name = paste0("S", 1:10), age = rnorm(10))
imp <- dmsa_import(B, pheno = ph)
imp
```

Three things happened, none silently. The probe axis was identified from the
dimension names and the matrix transposed; the value scale was detected
(betas here - M-values are converted with a note, because `dmsa_frame()`
converts betas to M-values itself and arriving twice-converted would corrupt
the frame); and the sheet was matched to the sample names by identifier.
Each of those is either unambiguous or an **error that says what to pass**:
`orientation =` when neither axis has probe-looking names, `values =` to
override the scale, `align_by =` to name the identifier column (minfi
`Basename` paths and `Sentrix_ID`/`Sentrix_Position` pairs are tried
automatically; `align_by = ".position"` declares same-order alignment
explicitly). A raw container - an `RGChannelSet`, a list of sesame `SigDF`s -
is refused with the preprocessing recipe rather than half-handled.

## The recipes

None of these packages are dmsa dependencies; the chunks are not run here.

```r
## minfi - any preprocessed object works via the SummarizedExperiment route
rg  <- minfi::read.metharray.exp(targets = targets)
gr  <- minfi::ratioConvert(minfi::mapToGenome(minfi::preprocessNoob(rg)))
imp <- dmsa_import(gr)                        # colData is the sheet
imp <- dmsa_import(minfi::preprocessRaw(rg))  # Meth/Unmeth handled too

## sesame - openSesame() returns the beta matrix directly
imp <- dmsa_import(sesame::openSesame(idat_dir), pheno = sheet)

## ewastools
mt  <- ewastools::read_idats(sheet$Basename)
imp <- dmsa_import(ewastools::dont_normalize(mt), pheno = sheet)

## meffil
qc   <- meffil::meffil.qc(samplesheet, verbose = FALSE)
norm <- meffil::meffil.normalize.quantiles(qc, number.pcs = 5)
imp  <- dmsa_import(meffil::meffil.normalize.samples(norm), pheno = samplesheet)

## ChAMP
load <- ChAMP::champ.load(directory, arraytype = "EPIC")
imp  <- dmsa_import(ChAMP::champ.norm(beta = load$beta), pheno = load$pd)

## RnBeads - extract, then import
imp <- dmsa_import(RnBeads::meth(rnb.set, row.names = TRUE),
                   pheno = RnBeads::pheno(rnb.set))

## methylprep - export betas to CSV, then import the file
imp <- dmsa_import("beta_values.csv", pheno = "sample_sheet.csv")
```

## Does the pipeline choice change the answer? A real-data check

If preprocessing choice changed DMSA's conclusions, the adapter would be a
convenience and a hazard. So we put one plate of real arrays through five
pipelines and asked the same two pre-registered questions of each result.

**Data.** 94 EPIC v1 arrays (one plate; 48 couples' families, 12 chips) from
Project Alpha's first batch - and, importantly, a *different physical scan*
of these participants than the arrays behind the package's bundled maps: the
two sets share zero barcodes. Person- and sample-level covariates (sex, age,
epithelial and fibroblast fractions) were joined from the canonical study
table by participant id; array-level structure (chip) entered as a random
intercept over this plate's own 12 chips; the canonical arrays'
control-probe surrogate variables were omitted, since they describe other
physical arrays. Across pipelines, everything was therefore identical except
the preprocessing.

**Questions.** (A) The pre-registered positive control: the immune,
inflammation & HLA system against the immune-cell fraction IC (with the
registered deviation - the epithelial fraction leaves the covariate set,
because IC = 1 - Epi - Fib makes the full set rank-deficient). Cell
composition is a signal known to dominate blood methylation; a method that
misses it is broken. (B) A theory pairing expected to stay quiet at this
sample size: the HPA axis against a symptom inventory (BSI). All runs:
`system > module > gene > probe`, B = 1999 permutations, seed 1.

**Result.** The control fired at the permutation floor in all five
pipelines - p = 0.0005 is 1/(B+1): not one of 1,999 permutations reached the
observed statistic - including on ewastools' completely unnormalized betas:

| pipeline    | preprocessing            | usable probes | system p | aligned direction |
|-------------|--------------------------|:---:|:------:|:---:|
| minfi noob  | noob background/dye      | 34  | 0.0005 | +   |
| minfi raw   | none (Meth/Unmeth ratio) | 35  | 0.0005 | +   |
| meffil      | functional normalization | 34  | 0.0005 | +   |
| sesame      | openSesame default       | 26  | 0.0005 | +   |
| ewastools   | none                     | 34  | 0.0005 | +   |

The agreement runs the whole cascade down. All 8 immune modules were
selected in every pipeline. At gene level, 18-20 genes were significant per
pipeline and **17 in all five** (Jaccard 0.81): the HLA class I/II battery
(HLA-DMB, -DOB, -DPA1, -DPB1, -DQA2, -DRA, -E, -F, HLA-F-AS1) with ICAM1,
IL10, IL12RB1, IL12RB2, IL18, NLRP3, TLR4 and TNF; gene-level aligned
directions were identical across all five pipelines for 92% of the shared
genes. And the null behaved: HPA x BSI came back non-significant in every
pipeline (p = 0.41-0.81), with the cascade never descending - zero modules
and zero genes selected, in any pipeline.

The one visible pipeline difference is coverage, not conclusions: sesame's
default masking left 26 usable probes where the others kept 34-35 - a
deliberate conservatism, reported by the frame, changing nothing downstream.

**Why this is expected, not lucky.** DMSA's families are declared in the map
rather than harvested from whatever the analyst loaded, the probe directions
are fixed by the direction catalogue before any data arrive, and the
permutation null is computed under the same transform as the observed
statistic. Preprocessing can move individual betas and drop probes; what it
does not get to do is redefine the hypothesis, the alignment, or the
reference distribution. The pipeline choice moves probe counts; the answer
stays the answer.

```r
## install
remotes::install_github("teindor/dmsa")
```

## Session info

```{r sessioninfo}
sessionInfo()
```
